---
title: "05_clean"
author: "Vinci Chan"
date: "`r Sys.Date()`"
format:
  html:
    toc: true
    toc_expand: true
    embed-resources: true
execute:
  warning: false
  message: false
---

# 1. Setup

```{r}
library(tidyverse)
library(arrow)
library(future)
library(furrr)

source("R/config.R")
source("R/utils_dates.R")

plan(multisession)

rquarters <- make_quarter_tbl(
  start_year = cfg$start_year,
  end_year = 2025,
  start_month_day = "0101",
  end_month_day = "1231"
) %>% 
mutate(
  rquarter = floor_date(date, "quarter") + months(3) - days(1)
  ) %>% 
filter(rquarter >= ymd(19990101)) %>% 
distinct(rquarter) %>% 
arrange(rquarter)

qa_dir_05 <- file.path(cfg$qa_dir, "05")
dir.create(qa_dir_05, recursive = TRUE, showWarnings = FALSE)
```

# 2. Functions

## 2.1. Extract rows and columns

```{r}
build_holdext_raw <- function() {
  all_path <- file.path(cfg$derived_dir, "all_holdings.parquet")
  extract_path <- file.path(cfg$derived_dir, "holdext_raw_parquet")

  if(!file.exists(extract_path)) {
    open_dataset(all_path) %>% 
    # Drop rows that are completely empty on key holding fields
    filter(
      !(
        is.na(nameOfIssuer) &
        is.na(titleOfClass) &
        is.na(cusip) &
        is.na(value) &
        is.na(shrsOrPrnAmt) &
        is.na(investmentDiscretion) &
        is.na(votingAuthority) &
        is.na(shares)
      )
    ) %>% 
    # Select only relevant rows
    select(
      filename, cik, cikname,
      rdate, fdate, nameOfIssuer,
      cusip, value, shares
    ) %>% 
    # Write holdext_raw
    write_dataset(
      extract_path,
      format = "parquet"
    )
  }
  open_dataset(extract_path) %>% 
    mutate(filename = cast(filename, string()))
}
```

## 2.2. Data Cleaning

### 2.2.1 rdate

We first clean and standardise the reporting and filing dates. `rdate` and `fdate` are converted from Unix time to calendar dates, obvious data‐entry errors in `rdate` are corrected manually, and all reports are mapped to canonical quarter‑end dates (`rquarter`) to simplify building manager-quarter panels and computing returns. To focus on the post‑EDGAR regime, we then drop all pre‑1999 reporting quarters.

```{r}
clean_dates <- function(holdext_raw) {
  # Change rdate and fdate class from Unix time to date
  holdext_raw <- holdext_raw %>% 
    mutate(
      rdate = as_date(rdate),
      fdate = as_date(fdate)
    )

  # Check for weird rdates
  weird_rdates <- holdext_raw %>% 
    to_duckdb() %>% 
    filter(rdate > fdate) %>% 
    select(filename, rdate, fdate) %>% 
    distinct(filename, .keep_all = TRUE) %>% 
    collect()

  holdext_raw <- holdext_raw %>% 
    mutate(
      # Fix weird dates manually
      rdate = case_when(
        filename == "0000080255-03-000400.txt" ~ as_date("2002-12-31"),
        filename == "0000905148-07-001531.txt" ~ as_date("2006-09-30"),
        filename == "0001103882-11-000001.txt" ~ as_date("2010-09-30"),
        TRUE ~ rdate
      ),
      # for filings with rdate == fdate, assume rdate to be the previous quarter end
      same_date = rdate == fdate,
      rdate = if_else(
        same_date,
        as_date(floor_date(fdate, unit = "quarter") - ddays(1)),
        rdate
      )
    ) %>%
    select(-same_date)

  # Standardise to rquarter and handle rdate == fdate
  rquarter_build <- holdext_raw %>% 
    distinct(rdate) %>% 
    collect() %>% 
    mutate(
      rquarter = floor_date(rdate, "quarter") + months(3) - days(1)
    )

  holdext_raw %>% 
    left_join(rquarter_build, by = "rdate") %>%
    relocate(rquarter, .after = rdate) %>%
    # Remove pre-1999 rquarters
    filter(rquarter >= ymd(19990101))
}
```

Weird `rdate`:
[0000080255-03-000400.txt](https://www.sec.gov/Archives/edgar/data/80255/0000080255-03-000400.txt)
[0000905148-07-001531.txt](https://www.sec.gov/Archives/edgar/data/1352851/0000905148-07-001531.txt)
[0001103882-11-000001.txt](https://www.sec.gov/Archives/edgar/data/1352910/0001103882-11-000001.txt)

### 2.2.2. fdate

Some rdates map to the same rquarter (quarter-end bucket), with multiple filings per manager-quarter. Some subsequent filings restate the complete, updated portfolio, while some are small additions that do not represent the full portfolio.

To obtain a single canonical portfolio per manager-quarter, for each `cik` + `rquarter` we first identify the filing(s) with the maximum number of holdings (n_holdings), as these are most likely to be full portfolio snapshots. If several filings have the same maximum n_holdings, we then select the one with the latest fdate as the final version for that reporting period. All non‑selected filings are dropped.

```{r}
select_canonical_filings <- function(holdext_raw) {
  canonical <- holdext_raw %>% 
  to_duckdb() %>% 
  group_by(cik, rquarter, filename) %>% 
  summarise(
    n_holdings = n(),
    fdate = max(fdate),
    .groups = "drop"
  ) %>% 
  group_by(cik, rquarter) %>% 
  filter(n_holdings == max(n_holdings)) %>% 
  slice_max(fdate, n = 1, with_ties = FALSE) %>% 
  ungroup() %>% 
  select(cik, rquarter, filename) %>% 
  collect()

  holdext_raw %>% 
    semi_join(canonical, by = join_by(cik, rquarter, filename))
}
```

### 2.2.3. cusip 

A few thousand of rows has cusips shorter than 7 characters. Those will be dropped.

```{r}
clean_cusip <- function(holdext_raw) {
  holdext_raw %>% 
    mutate(
      cusip = toupper(str_trim(cusip)),
      cusip_len = nchar(cusip)
    ) %>% 
    filter(cusip_len >= 7) %>% 
    select(-cusip_len)
}
```

There a few thousand rows with cusip as "000000000". They are kept for now.

### 2.2.4. value and shares outliers

```{r}
filter_outliers <- function(holdext_raw) {  
  caps <- holdext_raw %>%
    to_duckdb() %>% 
    summarise(
      value_cap = quantile(value, 0.999, na.rm = TRUE),
      shares_cap = quantile(shares, 0.999, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    collect()

  holdext_raw %>%
    filter(
      (is.na(value) | value <= caps$value_cap) & 
      (is.na(shares) | shares <= caps$shares_cap)
    )
}
```

### 2.2.5. value unit conversion

Starting on January 3, 2023, Form 13F filers report market value in dollars; previously, market value was reported in thousands of dollars. To maintain a consistent unit over time, we convert any dollar‑reported values back to thousands. [SEC Notice](https://www.sec.gov/files/form_13f_readme.pdf)

Rather than applying a blanket “post‑2023 = dollars” rule, we infer unit switches at the manager level using each filer’s own history:

For each CIK, we build a manager–quarter panel of aggregate portfolio value and holdings count, and classify the initial reporting regime using the first filing date (first_fdate): filers whose first 13F is before 2023‑01‑03 are treated as starting in thousands; those whose first 13F is on or after that date are treated as starting in dollars.

Within each manager’s panel, we compute quarter‑on‑quarter value ratios (value_ratio) and holdings ratios (holdings_ratio). We then identify at most one unit‑switch quarter per manager as the largest value spike with a large but plausible change in holdings (value_ratio > 50 and holdings_ratio < 5).

The manager’s unit state is modelled as a simple two‑state process (thousands vs dollars), starting from the initial regime and flipping once at the detected switch quarter, if any. All periods in the “dollars” state are then scaled down by dividing reported values by 1,000 at the individual‑holding level.

As a fallback, any manager whose aggregate portfolio value ever reaches 100 billion dollars in implied size is treated as reporting in dollars from that quarter onward, and its reported values are scaled down by 1,000.

In principle, we could add special rules for two remaining patterns: “dead” funds that suddenly revive after 2023 (E.g. CIK 0001727599), and pre‑2023 filers that quietly start reporting in dollars without a clean spike (E.g. CIK 0001544204). Handling these well would require a more complex multi‑episode state model (with switch‑on/switch‑off logic and extra heuristics), which I judge out of proportion to their impact. Instead, we accept that some of these edge cases may remain imperfectly classified and rely on the simple one‑switch rule plus the 100‑billion fallback to catch the most economically important mis‑scaled portfolios.

```{r}
# Heuristic calibration for unit-switch detection (not used downstream)
calibration_panel <- build_holdext_raw() %>%
  clean_dates() %>%
  select_canonical_filings() %>%
  clean_cusip() %>%
  filter_outliers() %>%
  to_duckdb() %>%
  group_by(cik, rquarter) %>%
  summarise(
    portfolio_value = sum(value, na.rm = TRUE),
    holdings_count  = n(),
    .groups         = "drop"
  ) %>%
  collect() %>%
  arrange(cik, rquarter) %>%
  group_by(cik) %>%
  mutate(
    lag_value    = lag(portfolio_value),
    lag_holdings = lag(holdings_count),
    value_ratio = if_else(
      !is.na(lag_value) & lag_value > 0,
      portfolio_value / lag_value,
      NA_real_
    ),
    holdings_ratio = if_else(
      !is.na(lag_holdings) & lag_holdings > 0,
      holdings_count / lag_holdings,
      NA_real_
    )
  ) %>%
  ungroup()

ratio_summaries <- calibration_panel %>%
  summarise(
    n_obs                = sum(!is.na(value_ratio)),
    p_value_gt_10        = mean(value_ratio > 10, na.rm = TRUE),
    p_value_gt_50        = mean(value_ratio > 50, na.rm = TRUE),
    p_holdings_gt_3      = mean(holdings_ratio > 3, na.rm = TRUE),
    p_holdings_gt_5      = mean(holdings_ratio > 5, na.rm = TRUE),
    p_spike_candidate    = mean(value_ratio > 50 & holdings_ratio < 5, na.rm = TRUE)
  )
ratio_summaries
```
```{r}
detect_unit_switches <- function(holdext_raw) {
  conversion_date     <- ymd("2023-01-03")
  conversion_rquarter <- floor_date(conversion_date, "quarter") + months(3) - days(1)

  # 1. Manager–quarter panel with simple ratios
  value_panel <- holdext_raw %>%
    to_duckdb() %>%
    group_by(cik, rquarter) %>%
    summarise(
      fdate           = max(fdate, na.rm = TRUE),
      portfolio_value = sum(value,  na.rm = TRUE),
      holdings_count  = n(),
      .groups         = "drop"
    ) %>%
    collect() %>%
    arrange(cik, rquarter) %>%
    group_by(cik) %>%
    mutate(
      first_rquarter = min(rquarter, na.rm = TRUE),
      first_fdate    = min(fdate,    na.rm = TRUE),

      # Use first_fdate, not rquarter
      initial_state  = if_else(first_fdate < conversion_date, 0L, 1L),

      lag_value    = lag(portfolio_value),
      lag_holdings = lag(holdings_count),

      value_ratio = if_else(
        !is.na(lag_value) & lag_value > 0,
        portfolio_value / lag_value,
        NA_real_
      ),
      holdings_ratio = if_else(
        !is.na(lag_holdings) & lag_holdings > 0,
        holdings_count / lag_holdings,
        NA_real_
      )
    ) %>%
    ungroup()

  # 2. For each CIK, pick the biggest spike with sane holdings
  switch_events <- value_panel %>%
    group_by(cik) %>%
    filter(
      !is.na(value_ratio),
      is.na(holdings_ratio) | holdings_ratio < 5
    ) %>%
    slice_max(value_ratio, n = 1, with_ties = FALSE) %>%
    mutate(
      switch_on_event = value_ratio > 50
    ) %>%
    select(cik, rquarter, switch_on_event)

  # 3. Build unit_state over time
  value_flags <- value_panel %>%
    left_join(switch_events, by = c("cik", "rquarter")) %>%
    group_by(cik) %>%
    mutate(
      switch_on_event = coalesce(switch_on_event, FALSE),
      event           = if_else(switch_on_event, 1L, 0L),
      unit_state      = initial_state + cumsum(event),
      unit_state      = as.integer(pmax(0, pmin(1, unit_state))),
      scale_down_flag = unit_state == 1L | cumsum(portfolio_value >= 1e8) > 0
    ) %>%
    ungroup() %>%
    select(cik, rquarter, value_ratio, holdings_ratio, scale_down_flag)

  # 4. Apply scaling back to row level
  holdext_raw %>%
    left_join(value_flags, by = c("cik", "rquarter"), copy = TRUE) %>%
    mutate(
      scale_down_flag = if_else(is.na(scale_down_flag), FALSE, scale_down_flag),
      value           = if_else(scale_down_flag, value / 1000, value)
    ) %>%
    select(
      filename, cik, cikname, rdate, rquarter, fdate,
      nameOfIssuer, cusip, value, shares
    )
}
```

### 2.2.6. Missing quarters

We build a complete cik-quarter panel with LOCF (last observation carried forward). However, only missing quarters between min_rquarter and max_rquarter will be filled. This keeps the data smooth while staying succinct. 

```{r}
build_missing_quarters_batch <- function(batch_ds) {
  # Identify the columns that define a holding row
  holding_cols <- c(
    "nameOfIssuer",
    "cusip",
    "value",
    "shares"
  )

  # 1. Load into memory
  batch_df <- batch_ds %>% collect()

  # 2. Complete grid
  cik_ranges <- batch_df %>% 
    distinct(cik, rquarter) %>% 
    group_by(cik) %>% 
    summarise(
      min_rquarter = min(rquarter),
      max_rquarter = max(rquarter),
      .groups = "drop"
    )
  
  cik_quarter_grid <- cik_ranges %>%
    cross_join(rquarters) %>%
    filter(
      rquarter >= min_rquarter,
      rquarter <= max_rquarter
    ) %>%
    select(cik, rquarter)

  filled_holdext_rquarter <- cik_quarter_grid %>% 
    left_join(batch_df, by = c("cik", "rquarter"))
  
  # 3. Flag missing vs non-operating quarters
  filled_holdext_rquarter <- filled_holdext_rquarter %>%
    mutate(
      is_empty_row = if_all(all_of(holding_cols), ~ is.na(.x)),
      missing_quarter = if_else(is_empty_row, 1L, 0L)
    ) %>%
    select(-is_empty_row)

  # 4. Compute prior_rdate for every row
  filled_with_prior <- filled_holdext_rquarter %>% 
    arrange(cik, rquarter) %>% 
    group_by(cik) %>% 
    mutate(
      own_rquarter = if_else(
        missing_quarter == 0L,
        rquarter,
        as.Date(NA)
      ),
      prior_rquarter_num = cummax(coalesce(as.numeric(own_rquarter), -Inf)),
      prior_rquarter = as_date(prior_rquarter_num)
    ) %>% 
    ungroup() %>% 
    select(-own_rquarter, -prior_rquarter_num)

  # 5. Build holdext_clean: keep real + non-OP rows, LOCF for missing
  real_holdings <- filled_with_prior %>% 
    filter(missing_quarter == 0L) %>% 
    select(cik:shares)

  core_cols <- colnames(real_holdings)

  locf_rows <- filled_with_prior %>% 
    filter(missing_quarter == 1L) %>% 
    select(cik, rquarter, prior_rquarter) %>% 
    left_join(
      real_holdings %>% 
        rename(prior_rquarter = rquarter),
      by = c("cik", "prior_rquarter"),
      relationship = "many-to-many"
    ) %>% 
    select(all_of(core_cols))

  bind_rows(real_holdings, locf_rows)
}
```

# 3. Orchestration

```{r}
# Stage 0
stage0 <- build_holdext_raw()

# Stage 1
stage1 <- stage0 %>%
  clean_dates() %>%
  select_canonical_filings() %>%
  clean_cusip() %>%
  filter_outliers() %>%
  detect_unit_switches()

# Stage 2
stage2_path <- file.path(cfg$staging_dir, "holdext_raw_stage2_parquet")
if (dir.exists(stage2_path)) unlink(stage2_path, recursive = TRUE, force = TRUE)

cik_batch <- stage1 %>%
  distinct(cik) %>%
  arrange(cik) %>%
  collect() %>%
  mutate(batch_id = sprintf("batch_%04d", ceiling(row_number() / 500)))

stage1 %>%
  left_join(cik_batch, by = "cik") %>%
  write_dataset(
    path = stage2_path,
    format = "parquet",
    partitioning = "batch_id"
  )

stage2 <- open_dataset(stage2_path)

# Stage 3
stage3_path <- file.path(cfg$derived_dir, "holdext_clean_parquet")
if (dir.exists(stage3_path)) unlink(stage3_path, recursive = TRUE, force = TRUE)
dir.create(stage3_path, recursive = TRUE)

batch_ids <- cik_batch %>%
  distinct(batch_id) %>%
  arrange(batch_id) %>%
  pull(batch_id)

future_walk(batch_ids, function(b) {
  batch_ds <- open_dataset(stage2_path) %>%
    filter(batch_id == b) %>%
    select(-batch_id)

  holdext_clean_batch <- build_missing_quarters_batch(batch_ds)

  write_parquet(
    holdext_clean_batch,
    sink = file.path(stage3_path, paste0(b, ".parquet")),
    compression = "zstd"
  )

  rm(holdext_clean_batch)
  gc()
  
  invisible(NULL)
})

holdext_clean <- open_dataset(stage3_path)
```

# 4. Check and clean up

```{r}
n_gaps <- holdext_clean %>%
  distinct(cik, rquarter) %>%
  collect() %>% 
  arrange(cik, rquarter) %>%
  group_by(cik) %>%
  mutate(
    expected_next = ceiling_date(rquarter, "month") %m+% months(3) - days(1),
    actual_next   = lead(rquarter),
    has_gap       = !is.na(actual_next) & actual_next != expected_next
  ) %>%
  ungroup() %>%
  filter(has_gap) %>%
  summarise(n_gap_rows = n()) %>%
  pull(n_gap_rows)

if(n_gaps == 0) {
  unlink(stage2_path, recursive = TRUE, force = TRUE)
}
```

# 5. QA

```{r}
holdext_raw   <- open_dataset(file.path(cfg$derived_dir,  "holdext_raw_parquet"))
holdext_clean <- open_dataset(file.path(cfg$derived_dir,  "holdext_clean_parquet"))
```

```{r}
quarter_cmp <- holdext_raw %>%
  select(cik, rdate) %>%
  collect() %>%
  mutate(
    rdate = as_date(rdate),
    rquarter = floor_date(rdate, "quarter") + months(3) - days(1)
  ) %>%
  distinct(cik, rquarter) %>%
  count(cik, name = "raw_n_q") %>%
  full_join(
    holdext_clean %>%
      distinct(cik, rquarter) %>%
      collect() %>%
      count(cik, name = "clean_n_q"),
    by = "cik"
  ) %>%
  mutate(
    raw_n_q   = replace_na(raw_n_q, 0L),
    clean_n_q = replace_na(clean_n_q, 0L),
    diff_q    = clean_n_q - raw_n_q
  )

p_quarters <- ggplot(quarter_cmp, aes(x = raw_n_q, y = clean_n_q)) +
  geom_abline(slope = 1, intercept = 0, color = "grey60", linetype = "dashed") +
  geom_point(aes(color = diff_q), alpha = 0.6, size = 1.2) +
  scale_color_gradient2(
    low = "#3b4cc0",
    mid = "#b0b0b0",
    high = "#f95d6a",
    midpoint = 0
  ) +
  labs(
    title = "Number of rquarters per CIK: holdext_raw vs holdext_clean",
    subtitle = "Points on the dashed line have equal quarter counts; points above it gained quarters in holdext_clean",
    x = "Raw number of rquarters",
    y = "Clean number of rquarters",
    color = "Clean - raw"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.margin = margin(t = 15, r = 15, b = 20, l = 15),
    plot.title.position = "plot",
    plot.subtitle = element_text(margin = margin(b = 10))
  )

ggsave(
  filename = file.path(qa_dir_05, "qa_quarters_raw_vs_clean.png"),
  plot     = p_quarters,
  width    = 8,
  height   = 5,
  dpi      = 300
)
```

```{r}
# Map raw rdate to quarter-end
raw_quarter_map <- holdext_raw %>%
  mutate(rdate = as_date(rdate)) %>%
  distinct(cik, rdate) %>%
  collect() %>% 
  mutate(
    rquarter = floor_date(rdate, "quarter") + months(3) - days(1)
  )

example_cik <- "0000002230"

raw_series <- holdext_raw %>%
  mutate(rdate = as_date(rdate)) %>%
  inner_join(raw_quarter_map, by = c("cik", "rdate")) %>%
  filter(cik == example_cik) %>%
  group_by(cik, rquarter) %>%
  summarise(raw_portfolio_value = sum(value, na.rm = TRUE), .groups = "drop")

clean_series <- holdext_clean %>%
  filter(cik == example_cik) %>%
  group_by(cik, rquarter) %>%
  summarise(clean_portfolio_value = sum(value, na.rm = TRUE), .groups = "drop")

both <- raw_series %>%
  inner_join(clean_series, by = c("cik", "rquarter")) %>%
  collect() %>%  
  mutate(
    raw_b   = raw_portfolio_value   / 1e9,
    clean_b = clean_portfolio_value / 1e9
  )

p_value <- ggplot(both, aes(x = rquarter)) +
  geom_line(aes(y = raw_b,   color = "Raw")) +
  geom_point(aes(y = raw_b,  color = "Raw"),  size = 1.2) +
  geom_line(aes(y = clean_b, color = "Clean")) +
  geom_point(aes(y = clean_b, color = "Clean"), size = 1.2) +
  labs(
    title = paste("Raw vs clean portfolio value for CIK", unique(both$cik)),
    x = NULL,
    y = "Portfolio value (billions USD)",
    color = NULL
  ) +
  theme_minimal(base_size = 12)

ggsave(
  filename = file.path(qa_dir_05, "qa_value_raw_vs_clean_example_cik.png"),
  plot     = p_value,
  width    = 7,
  height   = 4.5,
  dpi      = 300
)
```